\vspace{-0.5cm}The X-shooter GTO sample of GRB afterglow and host Galaxy spectra

### 2016-09-29 document/XSGRB_sample.tex
Adding GRB100316D

	\begin{abstract}
		Later
	\end{abstract}

@@

@@

	\input{tables/burst_overview.tex}

@@

@@

	\subsection{GRB100316D (z = 0.059)}
	The data presented here also formed the basis of GCN \#
	10512\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10512.gcn3}}, GCN \#
	10513\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10513.gcn3}}, GCN \#
	10543\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10543.gcn3}} and is published
	in \citet{Bufano2012} and \citet{Starling2011}. This GRB has an associated SN
	and has therefore undergone intense follow-up. The data presented here consists
	only of a subset of the entire VLT/X-shooter campaign. The transition from

>>>>

	\begin{abstract}
	The {\it Swift} satellite allows us to use gamma-ray bursts (GRBs) to peer into 
	the hearts of star forming galaxies through cosmic time. Our open
	collaboration, representing most of the active ESO member researchers in this
	field, seeks to build a public legacy sample of GRB X-shooter spectroscopy
	while {\it Swift} continues to fly. We propose to continue our programme to target
	all suitably observable GRB afterglows (up to 15 bursts per semester), with the
	primary goal of producing a well-defined, homogeneous, statistically useful
	sample. To date, our spectroscopy covers a redshift range from 0.059 to about
	8, with more than 20 robust metallicity measurements from absorption lines (over 
	the redshift range 1.7--5.9) and 4 secure detections of H$_2$ or CH molecular
	absorption. Such information is extremely difficult to obtain by other means.
	In terms of studying the spread and redshift evolution in gas-phase properties, the
	sample is still limited by low-number statistics. 
	\end{abstract}

@@

	Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
	towards about 10$^5$ QSOs have 5 systems with 
	$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$
	been identified (Noterdaeme et al.\ 2012, A\&A, 547, L1). Long GRB afterglow
	spectra, by contrast, reveal such systems in the majority of cases (e.g., Jakobsson et
	al.\ 2006, A\&A, 460, L13; Fynbo et al.\ 2009, ApJS, 185, 526).
	Whereas DLAs towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due
	to the atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing
	redshifts (e.g., Rafelski et al.\ 2014, ApJL, 782, L29), GRBs allow us to see 
	into the hearts of star-forming galaxies over
	the full history of cosmic star formation from $z \approx 0$ to $z > 8$ (e.g.,
	Tanvir et al.\ 2009, Nature, 461, 1254; Salvaterra et al.\ 2009, Nature, 461,
	1258; Jakobsson et al.\ 2012, ApJ, 752, 62). With afterglow spectroscopy
	(throughout the electromagnetic spectrum from X-rays to the sub-mm) we can hence
	characterize the properties of star-forming galaxies over cosmic history in
	terms of  redshifts, metallicities, molecular contents, ISM temperatures,
	UV-flux densities, etc.. This is, however, only possible as long as there are
	satellites in orbit that rapidly and accurately locate GRBs. The currently
	operating \textit{Swift} satellite, launched in 2004 and still fully-functioning,
	allows for very efficient follow-up observations of GRBs due to its
	unprecedented rate, speed and precision of localisations.

	%With GTO time on X-shooter we
	%have started building a sample of afterglows with X-shooter spectroscopy.
	%X-shooter is ideal for this with its high efficiency, wide wavelength coverage
	%and spectral resolution sufficient for inferring the fundamental physical
	%parameters just listed.

	\smallskip
	\smallskip

	{\bf The case for a large sample of GRBs with X-shooter spectroscopy.}

	There are
	several unanswered but fundamental questions that must be addressed in order
	to exploit the full potential of GRBs as cosmological probes.
	%Among them: what is the true redshift distribution and corresponding luminosity
	%function of long-duration GRBs?
	More than 50\% of the \textit{Swift} bursts with measured redshift are at $z >
	2$, and  5--10\% are expected to be above $z=5$ (Salvaterra et al. 2008, MNRAS, 380, L45; Salvaterra et al.\ 2012, ApJ, 749,
	68; Jakobsson et al.\ 2012, ApJ, 752, 62; Perley et al.\ 2016, ApJ, 817, 7). 
	A high redshift completeness is
	crucial for our understanding of the link between the number density of GRBs
	per unit redshift and the global star-formation history of the Universe, as
	measured by other means (UV, FIR, sub-mm, see Robertson \& Ellis 2011, ApJ, 744,
	95). The detection of GRBs at $z > 6$ shows that GRBs have become competitive as
	a tool to identifying galaxies at the highest redshifts and unsurpassed in
	providing detailed abundance information via absorption line spectroscopy
	(Tanvir et al.\ 2012, ApJ, 754, 46, McGuire et al.\ 2016, ApJ submitted, 
	arxiv:1512.07808).

	\smallskip

	%The sample of short GRBs with redshifts remains too small and too
	%incomplete to really
	%constrain their redshift distribution (Guetta \& Piran 2006, A\&A, 453, 823;
	%Salvaterra et al.\ 2008, MNRAS, 388, 6) making it of extreme importance to
	%enlarge as far as possible the number of known redshifts for this class of
	%sources.

	From March 2005 to March 2016 there have been about 350 {\it Swift} bursts 
	complying with
	our sample selection criteria (see ``Immediate objective''), and about half of them have
	measured redshifts. Among the latter subset, the team proposing these observations has
	measured about two thirds of the redshifts, mainly with FORS1/2 and X-shooter (see
	Fynbo et al.\ 2009, ApJS, 185, 526;  Jakobsson et al.\ 2012, ApJ, 752, 62;
	Kr\"uhler et al.\ 2012, ApJ, 758, 46). Our current aim is to build a sample superior to
	our previous low-resolution survey (Fynbo et al.\ 2009, ApJS, 185, 526), both in
	terms of quantity and  quality of the spectra. Our program started as 
	guaranteed time observations during
	periods 84-91, and we have continued in open time since then.

	%In P84 we started the X-shooter legacy survey in GTO time. Our aim is to  build
	%a sample of GRBs superior to our previous  low-resolution  survey, Fynbo et al.\
	%2009, ApJS, 185, 526) both in terms of quantity and  quality of the spectra. The
	%last period of GTO time on the survey was P90 (plus 6 hr of leftover GTO time
	%in  P91) and for P92 a proposal like this was  accepted for continuation of the
	%program in open time (concerning P93 see box 8).

	\smallskip

	X-shooter is in many ways
	the ideal GRB follow-up instrument and indeed GRB follow-up was one of the
	primary science cases behind the instrument design and implementation. Our
	program secures general purpose GRB afterglow spectroscopic follow-up 
	that adds strong legacy value to the \textit{Swift} GRB sample.
	Due to the wide wavelength range of X-shooter  
	with the same observation cover molecular H$_2$ absorption near the
	atmospheric cut-off and all the strong emission lines from the host in the
	NIR arm (e.g., Friis et al., 2015, MNRAS, 451, 167). In general, the wide wavelength coverage ensures that we always have
	features on which to base a redshift measurement as long as the afterglow is
	brighter than about 23 mag in either the $R$- or $z$-band. 
	Frequently, emission lines are also detected from the underlying host, which
	also provide further information such as SFR and metallicity (the top right 
	panel in Fig.~1 shows an example). Only for 
	7 out of more than 70 secured spectra could we not measure a redshift.   
	With the X-shooter survey we
	provide {\bf metallicity measurements} for about 30\% (Voigt-profile fits)
	of the $z>1.7$ events.
	So far we have
	measured metallicities for more than 20 GRB afterglows with X-shooter.
	With the wide wavelength coverage of X-shooter we can
	study important chemical species as Zn\,\textsc{ii}, Cr\,\textsc{ii} and $\alpha$
	elements over a much wider redshift range than what is possible with other
	instruments.
	As an example, we have measured a metallicity of $0.1 Z_\odot$ for GRB\,100219A
	at $z=4.669$ (Th{\"o}ne et al.\ 2013, MNRAS, 428, 3590), $0.02Z_\odot$ for 
	GRB\,111008A at $z=4.991$ (Sparre et al.\ 2014, ApJ, 785, 150) and $0.05Z_\odot$
	for the $z=5.9125$ GRB\,130606A (Hartoog et al.\ 2015, A\&A, 580, 139).  
	Reconciling the
	abundance patterns of GRB absorbers, other types of absorbers, QSO DLAs
	in particular, and old stars in the Local Group is an important long-term
	goal (see also Sparre et al.\ 2014, ApJ, 785, 150).
	Metallicities are also measured from host emission lines (Kr{\"u}hler et al.\ 2015,
	A\&A, 581, A125).
	GRB spectroscopy also allows us to determine the dust content of their environments,
	both through analysis of the depletion pattern and through measurement of the
	associated extinction (Japelj et al.\ 2015, A\&A, 451, 2050). 
	This allows us to quantify the dust-to-metals ratio and its
	evolution with redshift (e.g., De Cia et al.\ 2013, A\&A, 560, 88; Zafar \&
	Watson 2013, A\&A, 560, 26).
	%Moreover, GRB
	%afterglows allow investigation of random foreground absorbers, complementing
	%QSO sightline studies (e.g., Schulze et
	%al.\ 2012, A\&A, 546, 20).

	%In Fynbo et al. (2009, ApJS, 185,
	%526), we compare the host-galaxy HI
	%column density and metallicity (most of which has been measured by our group)
	%with a sample of damped \lya\ systems (DLAs) along QSO sightlines. We show that
	%most GRB-DLAs have a much larger column than QSO-DLAs.  The difference in
	%column density and metallicity for GRB and QSO sightlines shows that the total
	%cross-section for environments like those of GRBs must be less than a few
	%percent of the total cross-section for QSO-DLAs, and suggests that GRB
	%sightlines probe regions with high gas density.

	%The issue of production and {\bf release of Lyman continuum radiation} is
	%currently not fully understood, but the evidence suggests that star-forming
	%galaxies are dominating, in particular at the highest redshifts (e.g., Bianchi
	%et al.\ 2001, A\&A, 376, 1; Chen et al.\ 2007, ApJ, 667, L25). Also here
	%X-shooter spectroscopy will allow major progress due to the excellent blue
	%sensitivity.

	%GRB afterglows can probe dust in high-redshift star-forming galaxies
	%including exploring the {\bf 2175~\AA{} extinction bump}
	%(El\'iasd\'ottir et al.\ 2009, ApJ, 697, 1725; Zafar et al.\ 2011, A\&A,
	%732, 143).
	%Due to the broad wavelength  coverage we can do
	%this more accurately and over a broader range of redshifts with X-shooter.

	\smallskip

	We will also determine the frequency and properties of {\bf molecular
		absorption} towards GRB absorbers. Molecular gas is a key element to catalyze
	the process of star formation, but prior to our program H$_2$ had been detected
	just in two cases (tentatively in Fynbo et al.\ 2006, A\&A, 451, L47; securely
	in Prochaska et al.\ 2009, ApJ, 691, L27). With our X-shooter program we have
	found three more systems (Kr\"uhler et al.\ 2013, A\&A, 557, 18; D'Elia et al.\
	2014, A\&A, 564, 38; Friis et al.\, 2015, MNRAS, 451, 167). 
	We are currently analysing more of our spectra for less obvious molecular
	absorption and we expect to find more (a dedicated sample paper is 
	addressing this issue).

	\vspace{0.2cm}
	A natural question to ask is: {\bf how long should this work continue?} Our
	view is that we need to keep observing the afterglows as long as we have
	\textit{Swift} in operation. Also note that the program is still producing many 
	papers and provides data for many theses (Box 9 and 10). \textit{Swift} is currently 
	funded until 2018, but is likely to get more extensions given its overwhelming 
	success. As
	mentioned above GRBs allow us to probe star-forming galaxies that are almost
	impossible to study in other ways both in terms of redshifts, galaxy luminosity
	function, and regions within galaxies. After 7 periods we have 
	secured seven spectra of $z>4$ GRBs, of which three were of sufficient quality to
	allow abundance measurements (Th\"one et al.\ 2013, Sparre et al.\ 2014, 
	Hartoog et al.\ 2015). GRBs offer the only way to derive chemical abundances 
	for the gas phase of central, actively star-forming regions of high-$z$ galaxies. 
	The program also maintains a very high discovery
	potential where we occasionally find something completely unexpected that 
	provides interesting clues to puzzles in other fields, e.g. extinction of 
	type Ia supernovae (Fynbo et al.\ 2014, A\&A, 572, 12). Each of these spectra are like
	precious jewels – it is a type of observation that can never be repeated and a
	class of sightlines that can only be studied while we have operating GRB
	satellites.

	It is also worth adding that we have build up a rather unique team spread over
	Europe from Granada to Reykjavik, which by now has reached a point where the 
	distribution of night shifts, the scientific exploitation of the data is 
	efficient and where we are open to all new members who wish to participate.
	As mentioned all data are public immediately.

	{\bf For all of these reasons, we need to keep
		building up the sample of GRB afterglow spectra now as we may have to wait many
		years before a mission like \textit{Swift} becomes available again.}

	A significant proportion of GRBs lack a bright optical afterglow (``{\bf dark bursts}'', e.g.,
	Jakobsson et al.\ 2004, ApJ, 617, L21; Melandri et al.\ 2012, MNRAS, 421, 1265). 
	Some of these are at the highest
	redshifts ($z > 6$) and their observer-frame optical emission is absorbed by
	the IGM. The majority, however, suffer from large dust obscuration (e.g., Perley
	et al. 2009, AJ, 138, 1690; Greiner et al.\ 2011, A\&A, 526, 30).
	Identifying such GRBs is important for
	constraining the fraction of obscured star formation. In both cases,
	NIR emission is
	expected. X-shooter
	can adequately study these objects, provided that a NIR counterpart is timely
	identified, for which we have the dedicated HAWKI run D.

	The detection of {\bf absorption line variability} can reveal the burst
	influence on the surrounding medium and in turn the absorber distance from the
	burst and its metallicity (Vreeswijk et al.\ 2007, A\&A 468, 83; D'Elia et al.\
	2009, ApJ, 694, 332; Th\"one et al.\ 2011, MNRAS, 414, 479; De Cia et al.\
	2012, A\&A, 545, 64; Hartoog et al.\ 2013).
	{\bf Short GRBs} originate in a substantially different
	environment compared to long GRBs. Short GRBs may be related to the merging of
	compact binaries and the coalescence time can be long enough to allow the
	progenitor system to move far away from the star formation site (Belczynski et
	al.\ 2002, ApJ, 571, 147). Up to now, however, no spectrum with a sufficient 
	signal-to-noise ratio
	of a short GRB afterglow has been secured.
	A knowledge of the redshift distribution
	of short bursts is of key importance for the next generation of gravitational
	wave experiments, as they are the likely EM counterparts to their
	primary targets.

	\cleardoublepage

	\section{Sample selection criteria}

	Being of transient nature, it is difficult to impose strong sample selection
	criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
	restrictions exists from already, being it weather conditions, pointing
	restrictions of the telescope or poorly localized bursts as reported by the
	\textit{Swift}-telescope. To maximize the return of the follow-up campaign
	we have chosen a few selection criteria that attempts to provide an unbiased
	selection of bursts, while allowing for a high success-rate

	\footnote{Note that in the P84 proposal the criteria have
	been stated a bit differently, the visibility constraint being replaced by a
	declination + Sun angle constraint. The above criteria are however those
	defining the sample.}

	\begin{enumerate}
		\item GRB triggered onboard by Swift.
		\item Galactic $A_V \lesssim 0.5$ mag.
		\item XRT started observing within 10 minutes since the GRB; an XRT position must be distributed within 12 hr.
		\item The target must be visible for at least 60 min at least 30° above horizon, with the Sun below -12°.
		\item No bright closeby stars.

	\end{enumerate}

	\cleardoublepage

@@

@@

	\section{Reduction scheme}
	\input{tables/burst_overview.tex}

@@

	\subsection{GRB100316D (z = 0.059)}
	The data presented here also formed the basis of GCN \#
	10512\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10512.gcn3}}, GCN \#
	10513\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10513.gcn3}}, GCN \#
	10543\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10543.gcn3}} and is published
	in \citet{Bufano2012} and \citet{Starling2011}. This GRB is very close by and
	has an associated SN and has therefore undergone intense follow-up. The data
	presented here consists of a subset of the entire VLT/X-shooter campaign,
	covering the four first observing days while the afterglow still contributes
	significantly to the total emission. The first observations started 10 hours
	after the burst, before the SN was discovered, and targeted the star-forming
	'A'-region\citep{Starling2011}, not the GRB. A very rich spectrum containing a
	multitude of emission lines puts the host at $z = 0.059$. For three consequtive
	nights, 58, 79 and 101 hours after the \textit{Swift} trigger, the afterglow
	was observed as it transitioned into the spectrum of a high-velocity Ic-BL SN.
	The observations taken 79 and 101 hours after the burst are taken under
	programme 084.D-0265(A) (PI: Benetti), but with a similar setup to first two
	observations.
